Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-01 paper/semantics.tex
fix footnote mark spacing

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
To simplify the upcoming translation in \ref{sec:splitting}, 
we have chosen to statically distinguish between values and general computations,
using an underline constructor (``$\exv v$'') which explicitly note the parts of a computation that have been reduced to a value.
\footnote{In this presentation, the value/computation distinction is encoded intrinsically at the syntactic level.
However, we found it more convenient in the formal Twelf implementation to maintain the distinction with an extrinsic judgement.}
Moreover, the value/computation distinction interacts non-trivially with the three possible world classifications.
As a result, we end up with six classes of term in the grammar.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a reduction.
It also descends into {\tt gr} and {\tt next} blocks, 
using one of the other two step judgements ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, {\tt gr} blocks of ground values, and {\tt next} blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev{[\next e/x]e'}$, but this may potentially duplicate an
expensive computation $e$ depending on how many times $x$ appears in $e'$. 
Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}\end{abstrsyn}
This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
We read this as saying that somewhere within $e$ there was a subterm $q$ which needs to be hoisted out,
yielding the new term $e'$ which has a new variable $y$ where $q$ used to be.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

>>>>

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
To simplify the upcoming translation in \ref{sec:splitting}, 
we have chosen to statically distinguish between values and general computations,
using an underline constructor (``$\exv v$'') which explicitly note the parts of a computation that have been reduced to a value.%
\footnote{In this presentation, the value/computation distinction is encoded intrinsically at the syntactic level.
However, we found it more convenient in the formal Twelf implementation to maintain the distinction with an extrinsic judgement.}
Moreover, the value/computation distinction interacts non-trivially with the three possible world classifications.
As a result, we end up with six classes of term in the grammar.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a reduction.
It also descends into {\tt gr} and {\tt next} blocks, 
using one of the other two step judgements ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, {\tt gr} blocks of ground values, and {\tt next} blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev{[\next e/x]e'}$, but this may potentially duplicate an
expensive computation $e$ depending on how many times $x$ appears in $e'$. 
Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}\end{abstrsyn}
This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
We read this as saying that somewhere within $e$ there was a subterm $q$ which needs to be hoisted out,
yielding the new term $e'$ which has a new variable $y$ where $q$ used to be.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.%
\footnote{Because a program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

### 2016-01-01 paper/splitting.tex
fix footnote mark spacing

This structure can be described with a type; for instance, the staged quickselect yields a binary search tree.
Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs,
though we plan to pursue this in future work.
\footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 
replaced $\inl a$ and $\inr b$ with $\tup {0,a}$ and $\tup {1,b}$, and the proof of correctness would still have gone through.
We chose the tags, however, in order to keep the typed interpretation more natural, even in absence of a formal result.
}
\end{abstrsyn}

>>>>

This structure can be described with a type; for instance, the staged quickselect yields a binary search tree.
Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs,
though we plan to pursue this in future work.%
\footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 
replaced $\inl a$ and $\inr b$ with $\tup {0,a}$ and $\tup {1,b}$, and the proof of correctness would still have gone through.
We chose the tags, however, in order to keep the typed interpretation more natural, even in absence of a formal result.
}
\end{abstrsyn}
